Management induced changes of soil organic carbon on global croplands

### 2022-07-20 SOCbudgetPaper/discussion.Rmd
update to version before bug check

The substantial impact of changing management practices through time is indicated by the development of our estimated stock change factors (see Table \@ref(tab:SCFglo)) as well as by the time trend of the SOC debt (see Fig. \@ref(fig:SOCmaps)(a)). Residue management has changed over the last decades, especially with the phasing out of residue burning practices in several regions and increased general productivity, showing a clear impact on SOC dynamics --- underlining the importance to account for these effects in soil carbon modeling.

The Tier 2 approach [@ogle_cropland_in_ipcc_2019] used here is explicitly designed for agricultural soils, whereas we apply it also to soils under PNV. This is necessary in order to represent SOC losses under land-use change in a dynamic way, as this is an important driver of SOC dynamics. The comparison of simulated PNV data with LPJmL4 shows the model's capability in reproducing PNV SOC stocks (Fig. \@ref(fig:SOCLPJmL4)).
Concurrently, the point-data comparison (see Fig. \@ref(fig:SOCpoint)) shows low correlation for PNV, however also for cropland sites. This might also point to the fact that SOC stocks can vary at field and local scale considerable and thus a very high number of point data is needed to derive statistical properties, that could improve the point-to-grid-cell comparison [see @rammig_generic_2018].f

@@

In comparison with default stock change factors of the IPCC guidelines, our model estimates a stronger decline of SOC stocks (Table \@ref(tab:SCFglo)) for almost all regions. Tropical soils might suffer from low C input rates due to large yield gaps (Global Yield Gap and Water Productivity Atlas. Available URL: www.yieldgap.org (accessed on: 03/01/2022)) and high shares of residue removal and burning in lower-income countries [@smil_crop_1999; @williams_influence_1997; @jain_emission_2014]. Yet, even when comparing our estimates to the low-input stock change factors of the IPCC, our SOC loss is roughly twice as large as the revised 2019 IPCC default values [-@ogle_cropland_in_ipcc_2019], while it shows good agreement with the older default values from IPCC [-@lasco_cropland_in_ipcc_2006]. However, the revised estimates of the IPCC included much more and more recent data points, calling for a closer look on causes for the large deviations between our results and the refined Tier 1 factors. 
On the one hand, our approach does not account for unharvested carbon inputs from weeds or biomass cover on short-term fallows. Shifting agriculture with fallow periods might be prominent in tropical regions. While long-term fallow land (>4 years) is excluded from FAOSTAT as cropland, short-term fallow is not. Thus, our carbon inputs for these areas might be underestimated, leading to too low stock changes factors.
On the other hand, older studies by Don et al. [-@don_impact_2011] estimated SOC losses for tropical soils of around 25% on average corresponding to a stock change factor of 0.75, but also reported a wide range of measured SOC changes from -80% to +58%. Fujisaki et al. [-@fujisaki_forest_2015] however found much lower loss rates of around 9%, attributing the difference to the different time period length since the conversion to cropland. As our results do not specifically account for cropland age and most of the cropland is older than 20 years (as assumed for the default IPCC Tier 1 stock change factors) our stock change factors have to be lower by definition following the steady-state assumption that cropland will continue to approach a new equilibrium. 
For the same reason, our estimates for temperate regions might be lower than both IPCC [-@lasco_cropland_in_ipcc_2006] and IPCC [-@ogle_cropland_in_ipcc_2019] default values.
With the production-increasing impact of irrigation and fertilization on carbon-poor dryland soils, SOC under cropland can also be higher than under PNV with stock change factors above 1 (see Fig. \@ref(fig:SOCmaps)(d)), but these areas are much smaller than where the stock change factors are well below unity.

@@

The analysis of SOC stock gains and losses is complex and has several dimensions as climatic and anthropogenic effects overlap. There is broad consensus that land conversion to cropland has caused substantial C emissions over the historical period [e.g. @frielingstein_glob_carbon_2020]. There is uncertainty with respect to the overall size of these emissions from different methods and reference points and with respect to the contribution of cropland and agricultural management to these emissions. In order to mitigate greenhouse gas emissions, it is essential to stop the decline of SOC stocks or even transform cropland management to sequester atmospheric C in cropland soils [@minasny_soil_2017]. Defining the SOC debt of 1975 as the baseline, and measuring land-use emissions on cropland as the difference between a potential natural state and the state under human interventions [see @pugh_simulated_2015], we find that global cropland has acted as a emissions source since 1975. Annual C loss rates of 0.2 per 1000 C still have the opposite trend as the promoted 4 per 1000 C sequestration rate target [@minasny_soil_2017]. Dedicated efforts to increase cropland SOC are thus necessary, as management improvements at historical rates are not enough to counteract ongoing SOC degradation on cropland. Yet our study also shows the substantial impact of changing management on the development of SOC debt (Fig. \@ref(fig:SOCdebt) and Fig. \@ref(fig:SOCscen)).

>>>>

The substantial impact of changing management practices through time is indicated by the development of our estimated stock change factors (see Fig. \@ref(fig:SCFglo)) as well as by the time trend of the SOC debt (see Fig. \@ref(fig:SOCmaps)(a)). Residue management has changed over the last decades, especially with the phasing out of residue burning practices in several regions and increased general productivity, showing a clear impact on SOC dynamics --- underlining the importance to account for these effects in soil carbon modeling.

The Tier 2 approach [@ogle_cropland_in_ipcc_2019] used here is explicitly designed for agricultural soils, whereas we apply it also to soils under PNV. This is necessary in order to represent SOC losses under land-use change in a dynamic way, as this is an important driver of SOC dynamics. The comparison of simulated PNV data with LPJmL4 shows the model's capability in reproducing PNV SOC stocks (Fig. \@ref(fig:SOCLPJmL4)).
Concurrently, the point-data comparison (see Fig. \@ref(fig:SOCpoint)) shows low correlation for PNV, however also for cropland sites. This might also point to the fact that SOC stocks can vary at field and local scale considerably and thus a very high number of point data is needed to derive statistical properties, that could improve the point-to-grid-cell comparison [see @rammig_generic_2018].f

@@

In comparison with default stock change factors of the IPCC guidelines, our model estimates a stronger decline of SOC stocks (Fig. \@ref(fig:SCFglo)) for almost all regions. Tropical soils might suffer from low C input rates due to large yield gaps (Global Yield Gap and Water Productivity Atlas. Available URL: www.yieldgap.org (accessed on: 03/01/2022)) and high shares of residue removal and burning in lower-income countries [@smil_crop_1999; @williams_influence_1997; @jain_emission_2014]. Yet, even when comparing our estimates to the low-input stock change factors of the IPCC, our SOC loss is roughly twice as large as the revised 2019 IPCC default values [-@ogle_cropland_in_ipcc_2019], while it shows good agreement with the older default values from IPCC [-@lasco_cropland_in_ipcc_2006]. However, the revised estimates of the IPCC included much more and more recent data points, calling for a closer look on causes for the large deviations between our results and the refined Tier 1 factors. 
On the one hand, our approach does not account for unharvested carbon inputs from weeds or biomass cover on short-term fallows. Shifting agriculture with fallow periods might be prominent in tropical regions. While long-term fallow land (>4 years) is excluded from FAOSTAT as cropland, short-term fallow is not. Thus, our carbon inputs for these areas might be underestimated, leading to too low stock change factors.
On the other hand, older studies by Don et al. [-@don_impact_2011] estimated SOC losses for tropical soils of around 25% on average corresponding to a stock change factor of 0.75, but also reported a wide range of measured SOC changes from -80% to +58%. Fujisaki et al. [-@fujisaki_forest_2015] however found much lower loss rates of around 9%, attributing the difference to the different time period lengths since the conversion to cropland. As our results do not specifically account for cropland age and most of the cropland is older than 20 years (as assumed for the default IPCC Tier 1 stock change factors) our stock change factors have to be lower by definition following the steady-state assumption that cropland will continue to approach a new equilibrium. 
For the same reason, our estimates for temperate regions might be lower than both IPCC [-@lasco_cropland_in_ipcc_2006] and IPCC [-@ogle_cropland_in_ipcc_2019] default values.
With the production-increasing impact of irrigation and fertilization on carbon-poor dryland soils, SOC under cropland can also be higher than under PNV with stock change factors above 1 (see Fig. \@ref(fig:SOCmaps)(d)), but these areas are much smaller than where the stock change factors are well below unity.

@@

The analysis of SOC stock gains and losses is complex and has several dimensions as climatic and anthropogenic effects overlap. There is broad consensus that land conversion to cropland has caused substantial C emissions over the historical period [e.g. @frielingstein_glob_carbon_2020]. There is uncertainty with respect to the overall size of these emissions from different methods and reference points and with respect to the contribution of cropland and agricultural management to these emissions. In order to mitigate greenhouse gas emissions, it is essential to stop the decline of SOC stocks or even transform cropland management to sequester atmospheric C in cropland soils [@minasny_soil_2017]. Defining the SOC debt of 1975 as the baseline, and measuring land-use emissions on cropland as the difference between a potential natural state and the state under human interventions [see @pugh_simulated_2015], we find that global cropland has acted as a emissions source since 1975.
Comparing our SOC loss rate (the change of SOC debt) of $0.14\,\unit{GtC\,yr^{-1}}$ to estimates on land-use change induced emissions of $2.0\,\pm\,1.0\,\unit{GtC\,yr^{-1}}$ (sum of 'Bookkeeping LULCC emissions' and 'Loss of additional sink capacity' for the years 2009--2018 in Gasser et al. [-@gasser_historical_2020]), we find SOC emissions of the first 30 cm of the soil profile to be a minor contributor to overall land-use change induced emissions. 
Annual C loss rates of 0.2 per 1000 C still have the opposite trend as the promoted 4 per 1000 C sequestration rate target [@minasny_soil_2017]. Dedicated efforts to increase cropland SOC are thus necessary, as management improvements at historical rates are not enough to counteract ongoing SOC degradation on cropland. Yet our study also shows the substantial impact of changing management on the development of SOC debt (Fig. \@ref(fig:SOCdebt) and Fig. \@ref(fig:SOCscen)).

### 2022-07-20 SOCbudgetPaper/methods.Rmd
update to version before bug check

Firstly, in order to account for the impacts of legacy fluxes from land-use changes long before the time horizon of interest, we consider land-use change from 1510 onwards. In 1510, we assume all $SOC$ pools to be in natural steady-state, implying that all land-use change prior to that time occurs in 1510. We assume that by 1901, all cropland converted in 1510 has reached its new steady state, so that it is not necessary to explicitly account for even older land conversion. Model inputs for 1901--1930 for climate and natural vegetation litterfall are repeated for 1510--1900 to mimic constant climate conditions for this first initializing period. Similarly, agricultural management data are held constant at the level of 1965 until 1965. 
We acknowledge that this introduces a bias as agricultural management has changed prior to 1965, but this approach follows others studies looking on effects of land-use change and management [e.g. @schaphoff_lpjml4_2018-1; @herzfeld_soc_2021] and is limited by data availability on harvest statistics (and other management effects).

>>>>

Firstly, in order to account for the impacts of legacy fluxes from land-use changes long before the time horizon of interest, we consider land-use change from 1510 onwards. In 1510, we assume all $SOC$ pools to be in natural steady-state, implying that all land-use change prior to that time occurs in 1510. We assume that by 1901, all cropland converted in 1510 has reached its new steady state, so that it is not necessary to explicitly account for even older land conversion. Model inputs for 1901--1930 for climate and natural vegetation litterfall are repeated for 1510--1900 to mimic constant climate conditions for this first initializing period. Similarly, agricultural management data are held constant at the level of 1965 until 1965. 
We acknowledge that this introduces a bias as agricultural management has changed prior to 1965, but this approach follows others studies on effects of land-use change and management [e.g. @schaphoff_lpjml4_2018-1; @herzfeld_soc_2021] and is limited by data availability on harvest statistics (and other management effects).

### 2022-07-20 SOCbudgetPaper/results.Rmd
update to version before bug check

```{r captionSOCdebt, echo=FALSE}
captionSOCdebt <- "Global total $\\Delta SOC$ and $\\Delta SOC$ change for the first 30 cm of the soil profile. Panel (a) shows global $\\Delta SOC$ as the difference between $SOC$ under historical land use and $SOC$ under potential natural vegetation (PNV) in the year 2010 summed over all land-use types. Computing the difference between the $\\Delta SOC$ estimate for 2010 and for 1975 (b1) depicts areas of SOC depletion (SOC debt increase, red) and SOC accumulation (SOC debt decline, blue). Panel (b2) shows the of land-use change (LUC) induced change of ∆SOC between 1975 and 2010, whereas panel (b3) depicts the change due to changing agricultural management (MAN)."
```

```{r SOCdebt, out.width = "9cm", fig.cap = captionSOCdebt, echo=FALSE, fig.pos="h"}
# All defaults
knitr::include_graphics("../ResultNotebooks/Output/Images/SpatialDebtFigure.png")
```
The spatial distribution of the total $\Delta SOC$ summed over all land-use types (Fig. \@ref(fig:SOCdebt)(a))  and its change from 1975 to 2010 (Fig. \@ref(fig:SOCdebt)(b1)) reveals areas of SOC debt decline and increase. 
Regions with large cropland expansion (e.g. Brazil, Southeast Asia, Canada) continue to lose SOC, whereas regions with cropland reduction (and thus SOC restoration) or with accumulating cropland SOC can be found e.g. in highly productive areas of Europe and Central USA.

@@

To evaluate our modeled SOC stocks and stock changes under agricultural management, we compare our results to the default IPCC stock change factors $F^{\mathrm{SCF}}$ of 2006 [@lasco_cropland_in_ipcc_2006] and their refinements in 2019 [@ogle_cropland_in_ipcc_2019]. Both estimates are based on measurement data for cropland (see Table \@ref(tab:SCFglo)). To allow for comparison, we aggregate our stock change factors weighted by grid-level cropland area to derive average factors for the four IPCC climate zones (Fig. \@ref(fig:CLIMzone)). Note that IPCC Tier 1 factors are derived under the assumption that there is a linear change between steady states over 20 years, whereas our aggregated factors just reflect the relative change compared to a given potential natural vegetation reference stock without specifically tracking the time scale of the stock change.

```{r SCFglo, echo=FALSE, results = 'asis'}
table <- readRDS("../ResultNotebooks/Output/SFC_comparison.rds")
library(xtable)
print(xtable(table$comparison_table, digit=2, label="tab:SCFglo",
              caption="$F^{\\mathrm{SCF}}$ in comparison to IPCC Tier 1 default factors from the guidelines in 2006 \\citep{lasco_cropland_in_ipcc_2006} and the update in 2019 \\citep{ogle_cropland_in_ipcc_2019}."), comment=FALSE, caption.placement = "top")
```
Stock change factors for temperate climate zones of this study are lower than the default values of the IPCC. For the tropical regions the IPCC factors increased notably from the guidelines in 2006 [@lasco_cropland_in_ipcc_2006] to the update in 2019 [@ogle_cropland_in_ipcc_2019] due to the inclusion of more data points. Our results are in good agreement with the 2006 IPCC factors. Modeled $F^{SCF}$ have increased or stayed constant for all climate zones over time (1975-2010).

@@

```{r SOCglo, out.width = "11cm", fig.cap = captionSOCglo, echo=FALSE, fig.pos="h"}
# All defaults
knitr::include_graphics("../ResultNotebooks/Output/Images/glo_comparisonfigure.png")
```

>>>>

```{r captionSOCdebt, echo=FALSE}
captionSOCdebt <- "Global total $\\Delta SOC$ and $\\Delta SOC$ change for the first 30 cm of the soil profile. Panel (a) shows global $\\Delta SOC$ as the difference between $SOC$ under historical land use and $SOC$ under potential natural vegetation (PNV) in the year 2010 summed over all land-use types. Computing the difference between the $\\Delta SOC$ estimate for 2010 and for 1975 (b1) depicts areas of SOC depletion (SOC debt increase, red) and SOC accumulation (SOC debt decline, blue). Panel (b2) shows the of land-use change (LUC) induced change of $\\Delta SOC$ between 1975 and 2010, whereas panel (b3) depicts the change due to changing agricultural management (MAN)."
```

```{r SOCdebt, out.width = "6.3cm", fig.cap = captionSOCdebt, echo=FALSE, fig.pos="h"}
# All defaults
knitr::include_graphics("../ResultNotebooks/Output/Images/SpatialDebtFigure.png")
```
The spatial distribution of the total $\Delta SOC$ summed over all land-use types (Fig. \@ref(fig:SOCdebt)(a))  and its change from 1975 to 2010 (Fig. \@ref(fig:SOCdebt)(b1)) reveals areas of SOC debt decline and increase. 
Regions with large cropland expansion (e.g. Brazil, Southeast Asia, Canada) continue to lose SOC, whereas regions with cropland reduction (and thus SOC restoration) or with accumulating cropland SOC can be found e.g. in highly productive areas of Europe and Central USA.

@@

To evaluate our modeled SOC stocks and stock changes under agricultural management, we compare our results to the default IPCC stock change factors $F^{\mathrm{SCF}}$ of 2006 [@lasco_cropland_in_ipcc_2006] and their refinements in 2019 [@ogle_cropland_in_ipcc_2019]. Both estimates are based on measurement data for cropland (see Fig. \@ref(fig:SCFglo)). To allow for comparison, we aggregate our stock change factors weighted by grid-level cropland area to derive median factors for the four IPCC climate zones (Fig. \@ref(fig:CLIMzone)). Note that IPCC Tier 1 factors are derived under the assumption that there is a linear change between steady states over 20 years, whereas our aggregated factors just reflect the relative change compared to a given potential natural vegetation reference stock without specifically tracking the time scale of the stock change.

```{r captionSCFglo, echo=FALSE}
captionSCFglo <- "$F^{\\mathrm{SCF}}$ in comparison to IPCC Tier 1 default factors from the guidelines in 2006 \\citep{lasco_cropland_in_ipcc_2006} and the update in 2019 \\citep{ogle_cropland_in_ipcc_2019}. IPCC factors vary a bit with different types of management (low, medium, high). The results of the study span over a wide range -- indicated by the span between the 5th and 95th percentile -- but not overlapping with the IPCCs 2019 factors for tropical moist. The median for most regions is considerably lower than the IPCC factors."
```

```{r SCFglo, out.width = "15cm", fig.cap = captionSCFglo, echo=FALSE, fig.pos="h"}
# All defaults
knitr::include_graphics("../ResultNotebooks/Output/Images/SCF_comparison.png")
```

Stock change factors for temperate climate zones of this study are lower than the default values of the IPCC. For the tropical regions the IPCC factors increased notably from the guidelines in 2006 [@lasco_cropland_in_ipcc_2006] to the update in 2019 [@ogle_cropland_in_ipcc_2019] due to the inclusion of more data points. Our results span over a broad range due to the different ages of the cropland but also due to different agricultural management practices within climate regions.

@@

```{r SOCglo, out.width = "7cm", fig.cap = captionSOCglo, echo=FALSE, fig.pos="h"}
# All defaults
knitr::include_graphics("../ResultNotebooks/Output/Images/glo_comparisonfigure.png")
```
